

```{r setup}
library(Seurat)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(cowplot)
library(ggbreak) 
library(tidyr)
library(patchwork)
library(ComplexUpset)
library(ggrepel)
library(Polychrome)
library(ComplexHeatmap)
library(gplots)
library(RColorBrewer)
library(randomcoloR)

```

ln -s /home/common/data/output/projects/ENDO/E044/A033/EndoAtlas_GEO_meta_complete_downsampled.rds EndoAtlas.rds
EndoAtlas_meta <- readRDS(paste0(dir_data,"EndoAtlas_meta.rds"))
EndoAtlas <- readRDS(paste0(dir_data,"EndoAtlas.rds"))
EndoAtlas <- AddMetaData(EndoAtlas, EndoAtlas_meta)
saveRDS(EndoAtlas,paste0(dir_data,"EndoAtlas.rds") )

Add additional data to the Seurat metadata and reformat
```{r plotting of cell types per sample}
#Set directory links and create directories
dir_ENDO <- "/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/"

dir_data <- paste0(dir_ENDO,"_Data/")
dir_out <- paste0(dir_ENDO,"Figures_out/")
if (!dir.exists(dir_out)) {
  dir.create(dir_out, recursive = TRUE)
}

#Load data
EndoAtlas <- readRDS(paste0(dir_data, "EndoAtlas.rds"))
EndoAtlas_meta <- EndoAtlas@meta.data

#Preserve cell names
EndoAtlas_meta$cell_names <- rownames(EndoAtlas_meta)

##Modify menstrual cycle information
# Create column MenstrualCyclePhase_main based on MenstrualCyclePhase by merging the secretory sub-phases from 
EndoAtlas_meta$MenstrualCyclePhase_main <- as.character(EndoAtlas_meta$MenstrualCyclePhase)
EndoAtlas_meta$MenstrualCyclePhase_main[EndoAtlas_meta$MenstrualCyclePhase_main %in% c("secretory_early", "secretory_mid", "secretory_late")] <- "secretory"
#Factor samples by MenstrualCyclePhase_main for plotting
EndoAtlas_meta$MenstrualCyclePhase_main <- factor(EndoAtlas_meta$MenstrualCyclePhase_main, levels = c("proliferative","periovulatory","secretory"))

##Add additional metadata from manuscript Supplementary Table 1
# Read the tab-delimited text file
SupplementaryTable1 <- read.delim(paste0(dir_data, "SupplementaryTable1.txt"), header = TRUE, stringsAsFactors = FALSE)
# Perform a left join of the `meta` dataframe with `SupplementaryTable1`
EndoAtlas_meta <- dplyr::left_join(
  EndoAtlas_meta,
  SupplementaryTable1 %>% 
    dplyr::select(
      Patient.sample.ID,
      epithelial.pseudotime,
      sequencing.batch,
      X10x.capturing.batch,
      Minor.exclusion.criteria.logical,
      Menstrual.cycle.phase.marker.analysis,
      Cell.type.frequency.analysis..Figure.2e.f.,
      DEG.analysis..Figure.3a.b.,
      Ligand.receptor.analysis..Figure.3c.h.
    ),
  by = c("sample" = "Patient.sample.ID")
)

##Determine order of samples by MenstrualCyclePhase and epithelial.pseudotime
MenstrualCyclePhase_sampleOrder <- EndoAtlas_meta%>%
  select(sample, MenstrualCyclePhase,epithelial.pseudotime) %>%
  distinct() %>%
  arrange(MenstrualCyclePhase, epithelial.pseudotime) %>%
  mutate(MenstrualCyclePhase_sampleOrder = row_number()) %>%
  ungroup()
#Factor samples by MenstrualCyclePhase and epithelial.pseudotime for plotting
EndoAtlas_meta$sample <- factor(EndoAtlas_meta$sample, levels = c(MenstrualCyclePhase_sampleOrder$sample))

#Add cell names as row names
rownames(EndoAtlas_meta) <- EndoAtlas_meta$cell_names

#Save EndoAtlas_meta as .rds
saveRDS(EndoAtlas_meta, paste0(dir_data, "EndoAtlas_meta.rds"))

#EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))

```
